---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(citation.bibtex.max = 0)
```

# BGmisc

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R package version](https://www.r-pkg.org/badges/version/ggBGmisc)](https://cran.r-project.org/package=ggBGmisc)
[![Package downloads](https://cranlogs.r-pkg.org/badges/grand-total/ggBGmisc)](https://cran.r-project.org/package=BGmisc)</br>
[![R-CMD-check](https://github.com/R-Computing-Lab/ggBGmisc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/R-Computing-Lab/ggBGmisc/actions/workflows/R-CMD-check.yaml)
[![Dev Main
branch](https://github.com/R-Computing-Lab/ggBGmisc/actions/workflows/R-CMD-dev_maincheck.yaml/badge.svg)](https://github.com/R-Computing-Lab/ggBGmisc/actions/workflows/R-CMD-dev_maincheck.yaml)
[![codecov](https://codecov.io/gh/R-Computing-Lab/ggBGmisc/graph/badge.svg)](https://codecov.io/gh/R-Computing-Lab/ggBGmisc)
![License](https://img.shields.io/badge/License-GPL_v3-blue.svg)
<!-- badges: end -->

The ggBGmisc R package offers a comprehensive suite of functions tailored for extended behavior genetics analysis, including model identification, calculating relatedness, pedigree conversion, pedigree simulation, and more.

## Installation

You can install the released version of BGmisc from [CRAN](https://cran.r-project.org/) with:

```r
install.packages("ggBGmisc")
```

To install the development version of BGmisc from [GitHub](https://github.com/) use:

```r
# install.packages("devtools")
devtools::install_github("R-Computing-Lab/ggBGmisc")
```

## Citation

If you use ggBGmisc in your research or wish to refer to it, please cite the following paper:

```
citation(package = "ggBGmisc")
```
```{r,comment='',results='asis',echo=FALSE} 
print(citation(package = "ggBGmisc"), style = "text", bibtex = FALSE)
```

A BibTeX entry for LaTeX users is
```{r,comment='',echo=FALSE}
toBibtex(citation(package = "ggBGmisc"))
```

## Contributing
Contributions to the ggBGmisc project are welcome. For guidelines on how to contribute, please refer to the [Contributing Guidelines](https://github.com/R-Computing-Lab/ggBGmisc/blob/main/CONTRIBUTING.md). Issues and pull requests should be submitted on the GitHub repository. For support, please use the GitHub issues page.


### Branching and Versioning System
The development of ggBGmisc follows a [GitFlow branching strategy](https://tilburgsciencehub.com/topics/automation/version-control/advanced-git/git-branching-strategies/):

- **Feature Branches**: All major changes and new features should be developed on separate branches created from the dev_main branch. Name these branches according to the feature or change they are meant to address.
- **Development Branches**: Our approach includes two development branches, each serving distinct roles:
  - **`dev`**: This branch is the final integration stage before changes are merged into the `main` branch. It is considered stable, and only well-tested features and updates that are ready for the next release cycle are merged here.
- **Main Branch** (`main`): The main branch mirrors the stable state of the project as seen on CRAN. Only fully tested and approved changes from the dev_main branch are merged into main to prepare for a new release.


## License
ggBGmisc is licensed under the GNU General Public License v3.0. For more details, see the [LICENSE.md](https://github.com/R-Computing-Lab/BGmisc/blob/main/LICENSE.md) file.

