---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(citation.bibtex.max = 0)
```

# ggpedigree

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R package version](https://www.r-pkg.org/badges/version/ggpedigree)](https://www.r-pkg.org/badges/version/ggpedigree)
[![Package downloads](https://cranlogs.r-pkg.org/badges/grand-total/ggpedigree)](https://cranlogs.r-pkg.org/badges/grand-total/ggpedigree)</br>
[![R-CMD-check](https://github.com/R-Computing-Lab/ggpedigree/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/R-Computing-Lab/ggpedigree/actions/workflows/R-CMD-check.yaml)
[![Dev branch](https://github.com/R-Computing-Lab/ggpedigree/actions/workflows/R-CMD-devcheck.yaml/badge.svg)](https://github.com/R-Computing-Lab/ggpedigree/actions/workflows/R-CMD-devcheck.yaml)
[![codecov](https://codecov.io/gh/R-Computing-Lab/ggpedigree/branch/main/graph/badge.svg?token=xXWYDcD9CF)](https://codecov.io/gh/R-Computing-Lab/ggpedigree)
![License](https://img.shields.io/badge/License-GPL_v3-blue.svg)
<!-- badges: end -->

The ggpedigree R package complements the BGmisc package by providing a set of functions for visualizing pedigree data using ggplot2 and plotly. It is designed to work seamlessly with the BGmisc package, which provides tools for simulating, manipulating, and modeling pedigree data. It extends the plotting functionality of the `kinship2` package, which is widely used for pedigree analysis in R.

## Installation

You can install the released version of ggpedigree from [CRAN](https://cran.r-project.org/) with:

```r
install.packages("ggpedigree")
```

To install the development version of ggpedigree from [GitHub](https://github.com/) use:

```r
# install.packages("devtools")
devtools::install_github("R-Computing-Lab/ggpedigree")
```

## Citation

If you use ggpedigree in your research or wish to refer to it, please cite the following:

```
citation(package = "ggpedigree")
```

```{r,comment='',results='asis',echo=FALSE} 
print(citation(package = "ggpedigree"), style = "text", bibtex = FALSE)
```

A BibTeX entry for LaTeX users is
```{r,comment='',echo=FALSE}
toBibtex(citation(package = "ggpedigree"))
```

## Contributing
Contributions to the ggpedigree project are welcome. For guidelines on how to contribute, please refer to the [Contributing Guidelines](https://github.com/R-Computing-Lab/ggpedigree/blob/main/CONTRIBUTING.md). Issues and pull requests should be submitted on the GitHub repository. For support, please use the GitHub issues page.


### Branching and Versioning System
The development of ggpedigree follows a [GitFlow branching strategy](https://tilburgsciencehub.com/topics/automation/version-control/advanced-git/git-branching-strategies/):

- **Feature Branches**: All major changes and new features should be developed on separate branches created from the dev branch. Name these branches according to the feature or change they are meant to address.
- **Development Branch**: Our approach includes two development branches, each serving distinct roles:
  - **`dev`**: This branch is the final integration stage before changes are merged into the `main` branch. It is considered stable, and only well-tested features and updates that are ready for the next release cycle are merged here.
- **Main Branch** (`main`): The main branch mirrors the stable state of the project as seen on CRAN. Only fully tested and approved changes from the dev_main branch are merged into main to prepare for a new release.


## License
ggpedigree is licensed under the GNU General Public License v3.0. For more details, see the [LICENSE.md](https://github.com/R-Computing-Lab/ggpedigree/blob/main/LICENSE.md) file.

