---
title: "Plotting pedigrees with `ggPedigree()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting pedigrees with `ggPedigree()`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```



# Introduction

This vignette demonstrates pedigree visualization with `ggPedigree()` from `ggpedigree`. Pedigree plots are central to fields like human genetics, behavioral science, and genealogy, where understanding lineage and familial relationships informs diagnosis, analysis, and interpretation. 

`ggPedigree()` builds on `ggplot2` and `kinship2`, leveraging modular helpers from both `{BGmisc}` and `{ggpedigree}` to streamline plot generation while retaining full aesthetic flexibility.

Unlike base R pedigree tools, this package returns a `ggplot2` object, meaning users can customize their output using familiar `ggplot2` syntax, themes, scales, and layering. Throughout this vignette, we walk through basic usage, layout control, status overlays, aesthetic customization, and faceting for multifamily visualization.

We use two bundled example datasets (from `BGmisc`) to illustrate the package's capabilities. The first dataset, `potter`, contains a fictional wizarding family tree, while the second dataset, `hazard`, includes a a multigenerational historical pedigree with affected/unaffected status. 


# Basic usage

We begin by loading the required libraries. These include the main plotting package `ggpedigree`, helper utilities from `BGmisc`, and supporting tools for aesthetic and data manipulation.

```{r libraries, message=FALSE, warning=FALSE}
library(ggpedigree) # ggPedigree lives here
library(BGmisc) # helper utilities & example data
library(ggplot2) # ggplot2 for plotting
library(viridis) # viridis for color palettes
library(tidyverse) # for data wrangling
```


The simplest usage requires a data frame and column names for family and individual IDs. Here’s a basic pedigree plot using the `potter` data:

```{r basic-usage}
data("potter")
ggPedigree(potter,
  famID = "famID",
  personID = "personID"
)
```

Behind the scenes, `ggPedigree()`:

1. reshapes the data by family (`ped2fam()`),

2. recodes sex (e.g., 0/1/NA) into semantic labels via `recodeSex()`

3. extracts a layout grid with `calculateCoordinates()`

4. builds connection segments for spouses, siblings, parents, and offspring with `calculateConnections()`

The result is returned as a `ggplot2` object, which allows immediate post-processing with standard `ggplot` syntax.



# Customizing Aesthetics

Users can control aesthetics via the `config` list. 
Most appearance settings are managed through a `config` list. If any options are omitted, they fall back to predefined defaults. For example, we can turn off sex-based fill color and assign specific colors to connection segments:

```{r customize-aesthetics}
ggPedigree(
  potter,
  famID = "famID",
  personID = "personID",
  config = list(
    code_male = 1,
    sex_color = FALSE,
    spouse_segment_color = "pink",
    sibling_segment_color = "blue",
    parent_segment_color = "green",
    offspring_segment_color = "black"
  )
)
```

As with any `ggplot2` object, further theming and labeling can be applied using layers or themes. For instance:


```{r}
ggPedigree(potter,
  famID = "famID",
  personID = "personID"
) +
  theme_bw(base_size = 12)
```
This can be particularly useful for matching house styles in academic publications or removing clutter for presentations.



# Overlaying Status Information

In many applied settings, pedigrees include binary phenotypes such as affected/unaffected status. To incorporate this, use the `status_col` argument to point to a column containing the relevant variable.

```{r}
data("hazard")

p <- ggPedigree(
  hazard,
  famID = "famID",
  personID = "ID",
  status_col = "affected",
  config = list(
    code_male = 0,
    sex_color = TRUE,
    affected = TRUE,
    unaffected = FALSE,
    affected_shape = 4
  )
)

p
```

The function will automatically assign different shapes to affected individuals. If the `status_col` isn't already a factor, `ggPedigree()` will coerce it. You can explicitly set what values correspond to "affected" and "unaffected" via the `config` list.

If you turn off sex-based coloring, `ggPedigree()` will apply default `ggplot2` fill scales based on status:


You can also add affected/unaffected status to the plot. The `status_col` argument allows you to specify a column in your data frame that contains the affected/unaffected status of each individual. This column should be a factor or character vector, and `ggPedigree()` will automatically recode it to a factor if necessary.

When you add the `status_col` argument, `ggPedigree()` will automatically color the affected individuals in one color and the unaffected individuals in another. You can indicate what the values of the affected/unaffected status are in your data frame using the `affected` and `unaffected` arguments in the `config` list. By default, `ggPedigree()` will overlay the affected individuals with a different shape, but you can customize this as well.

```{r}
data("hazard")

p <- ggPedigree(
  hazard, # %>% mutate(affected = as.factor(ifelse(affected == TRUE, "affected", "uneffected"))),
  famID = "famID",
  personID = "ID",
  status_col = "affected",
  config = list(
    code_male = 0,
    sex_color = TRUE,
    affected = TRUE,
    unaffected = FALSE,
    affected_shape = 4
  )
)

p
```

If you set sex_color to FALSE, the affected values will be filled with the default color palette.

```{r}
ggPedigree(
  hazard,
  famID = "famID",
  personID = "ID",
  status_col = "affected",
  config = list(
    code_male = 0,
    sex_color = FALSE,
    affected = TRUE,
    unaffected = FALSE
  )
)
```

# Multiple families in one graphic

If you have multiple families in your data frame, you can use the `facet_wrap()` function from `ggplot2` to create separate plots for each family. This is useful if you want to compare the pedigrees of different families side by side.

Facet to separate distinct pedigrees, note that you can use `scales = "free_x"` to allow each family to have its own x-axis scale.

```{r}
p +
  facet_wrap(~famID, scales = "free_x")
```

# Changing the layout

```{r}
p +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line        = element_line(colour = "black"),
    axis.text.x      = element_blank(),
    axis.text.y      = element_blank(),
    axis.ticks.x     = element_blank(),
    axis.ticks.y     = element_blank(),
    axis.title.x     = element_blank(),
    axis.title.y     = element_blank()
  ) + scale_color_viridis(
    discrete = TRUE,
    labels = c("Female", "Male", "Unknown")
  )
```

# More complex examples

```{r message=FALSE, warning=FALSE}
library(BGmisc) # helper utilities & example data

data("inbreeding")

df <- inbreeding

# df  <- dplyr::filter(df, famID %in% c(5, 7))


p <- ggPedigree(
  df,
  famID = "famID",
  personID = "ID",
  status_col = "proband",
  #  debug = TRUE,
  config = list(
    code_male = 0,
    sex_color = F,
    #  label_method = "geom_text",
    affected = TRUE,
    unaffected = FALSE,
    generation_height = 2,
    generation_width = 1,
    affected_shape = 4,
    spouse_segment_color = "pink",
    sibling_segment_color = "blue",
    parent_segment_color = "green",
    offspring_segment_color = "black"
  )
)

# p$connections%>%filter(personID     ==60) %>% nrow()
# p$connections%>%filter(personID     ==66) %>% unique()
# p$connections%>%filter(personID     ==65) %>% unique()

# p$connections%>%filter(personID  >=61 &
#       personID  <62 ) %>% unique()

p + facet_wrap(~famID, scales = "free") #+ scale_color_viridis(
#   discrete = TRUE,
#   labels = c("TRUE", "FALSE")
#  )  + theme_bw(base_size = 14)  +  guides(colour="none", shape="none")
```

# Power paper

```{r}
library(tibble)

pedigree_df <- tribble(
  ~personID, ~momID, ~dadID, ~sex, ~famID,
  10011, NA, NA, 0, 1,
  10012, NA, NA, 1, 1,
  10021, NA, NA, 1, 1,
  10022, 10011, 10012, 1, 1,
  10023, 10011, 10012, 0, 1,
  10024, NA, NA, 0, 1,
  10025, NA, NA, 0, 1,
  10026, 10011, 10012, 0, 1,
  10027, 10011, 10012, 1, 1,
  10031, 10023, 10021, 0, 1,
  10032, 10023, 10021, 1, 1,
  10033, 10023, 10021, 1, 1,
  10034, 10023, 10021, 1, 1,
  10035, 10023, 10021, 0, 1,
  10036, 10024, 10022, 1, 1,
  10037, 10024, 10022, 0, 1,
  10038, 10025, 10027, 1, 1,
  10039, 10025, 10027, 0, 1,
  10310, 10025, 10027, 1, 1,
  10311, 10025, 10027, 1, 1,
  10312, 10025, 10027, 0, 1,
  10011, NA, NA, 0, 2,
  10012, NA, NA, 1, 2,
  10021, NA, NA, 0, 2,
  10022, 10011, 10012, 0, 2,
  10023, 10011, 10012, 1, 2,
  10024, 10011, 10012, 1, 2,
  10025, NA, NA, 1, 2,
  10026, 10011, 10012, 0, 2,
  10027, NA, NA, 1, 2,
  10031, 10021, 10023, 1, 2,
  10032, 10021, 10023, 0, 2,
  10033, 10021, 10023, 1, 2,
  10034, 10022, 10025, 0, 2,
  10035, 10022, 10025, 0, 2,
  10036, 10022, 10025, 1, 2,
  100310, 10022, 10025, 1, 2,
  10037, 10026, 10027, 0, 2,
  10038, 10026, 10027, 0, 2,
  10039, 10026, 10027, 0, 2,
  100311, 10026, 10027, 1, 2,
  100312, 10026, 10027, 1, 2
) %>% mutate(proband = TRUE)

# pedigree_df <- recodeSex(pedigree_df,code_male = 1, recode_male = "M")
pedigree_df$personID[pedigree_df$famID == 1] <- pedigree_df$personID[pedigree_df$famID == 1] - 10000
pedigree_df$momID[pedigree_df$famID == 1] <- pedigree_df$momID[pedigree_df$famID == 1] - 10000
pedigree_df$dadID[pedigree_df$famID == 1] <- pedigree_df$dadID[pedigree_df$famID == 1] - 10000



p <- ggPedigree(
  pedigree_df,
  famID = "famID",
  personID = "personID",
  status_col = "proband",
  #  debug = TRUE,
  config = list(
    code_male = 1,
    sex_color = F,
    apply_default_scales = FALSE,
    label_method = "geom_text",
    #   affected = TRUE,
    #   unaffected = FALSE,
    generation_height = 1,
    generation_width = 1,
    affected_shape = 4,
    spouse_segment_color = "black",
    sibling_segment_color = "black",
    parent_segment_color = "black",
    offspring_segment_color = "black"
  )
)

p + scale_shape_manual(
  values = c(16, 15, 15),
  labels = c("Female", "Male", "Unknown")
) +
  guides(colour = "none", shape = "none")
# guides(colour="none", shape="none") +
# facet_wrap(~famID, scales= "free")
```
