---
title: "Plotting pedigrees with `ggPedigree()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting pedigrees with `ggPedigree()`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```



# Introduction

This vignette demonstrates pedigree visualisation with `ggPedigree()` from `ggpedigree`.
Internally, `ggPedigree()` leverages helpers from both {BGmisc} (ped2fam(), recodeSex()) and {ggpedigree} (`calculateCoordinates()`, `calculateConnections()`), and returns a standard `ggplot2` object, so any familiar ggplot syntax can be added afterward.

We illustrate usage with two bundled data sets:

- `potter` – extended wizarding family

- `hazard` – historical royalty


# Basic usage

```{r libraries, message=FALSE, warning=FALSE}
library(ggpedigree) # ggPedigree lives here
library(BGmisc) # helper utilities & example data
library(ggplot2) # ggplot2 for plotting
library(viridis) # viridis for color palettes
library(tidyverse) # for data wrangling
```

```{r}
data("potter")
ggPedigree(potter,
  famID_col = "famID",
  personID_col = "personID",
  code_male = 1
)
```

`ggPedigree()` automatically:

1. reshapes the data by family (`ped2fam()`),

2. recodes sex (`recodeSex()`),

3. computes node layout (`calculateCoordinates()`), and

4. draws spouse, parent, sibling, and offspring links (`calculateConnections()`).

The result is a `ggplot2` object, so you can add any ggplot syntax to it. 


# Customizing the plot

All aesthetics reside in a config list; anything you omit falls back to the default.

```{r}
ggPedigree(
  potter,
  famID_col = "famID",
  personID_col = "personID",
  code_male = 1,
  config = list(
    spouse_segment_color = "pink",
    sibling_segment_color = "blue",
    parent_segment_color = "green",
    offspring_segment_color = "black"
  )
)
```

Because the result is just a ggplot object, regular layering applies:

```{r}
ggPedigree(potter,
  famID_col = "famID",
  personID_col = "personID",
  code_male = 1
) +
  theme_bw(base_size = 12)
```



# Alternative coding schemes

Pass the appropriate `code_male` when males are coded differently.


```{r}
data("hazard")

p <- ggPedigree(
  hazard, #%>% mutate(affected = as.factor(ifelse(affected == TRUE, "affected", "uneffected"))),
  famID_col = "famID",
  personID_col = "ID",
  code_male = 0,
  status_col = "affected",
  config =list (affected=TRUE,
                unaffected=FALSE,
                affected_shape = 4
))

p
```

# Multiple families in one graphic

Facet to separate distinct pedigrees:
```{r}
p +
  facet_wrap(~famID, scales = "free_x")
```

# Changing the layout

```{r}
p +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line        = element_line(colour = "black"),
    axis.text.x      = element_blank(),
    axis.text.y      = element_blank(),
    axis.ticks.x     = element_blank(),
    axis.ticks.y     = element_blank(),
    axis.title.x     = element_blank(),
    axis.title.y     = element_blank()
  ) + scale_color_viridis(
    discrete = TRUE,
    labels = c("Female", "Male", "Unknown")
  )
```
