---
title: "Plotting more complicated pedigrees with `ggPedigree()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting pedigrees with `ggPedigree()`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```



# Introduction

This vignette demonstrates pedigree visualization with `ggPedigree()` from `ggBGmisc`.
Internally, `ggPedigree()` leverages helpers from both {BGmisc} (ped2fam(), recodeSex()) and {ggBGmisc} (`calculateCoordinates()`, `calculateConnections()`), and returns a standard `ggplot2` object, so any familiar ggplot syntax can be added afterward.

We illustrate usage with a more complicated data set:

- `asoiaf` – extended Targaryen family


# Basic usage

```{r libraries, message=FALSE, warning=FALSE}
library(ggpedigree) # ggPedigree lives here
library(BGmisc) # helper utilities & example data
library(ggplot2) # ggplot2 for plotting
library(viridis) # viridis for color palettes
library(tidyverse) # for data wrangling
```



## Load Packages and Data

We begin by loading the required libraries and examining the structure of the built-in `ASOIAF` pedigree.

```{r echo=TRUE, message=FALSE, warning=FALSE}

data(ASOIAF)
```

The ASOIAF dataset includes character IDs, names, family identifiers, and  parent identifiers for a subset of characters drawn from the *A Song of Ice and Fire* canon.

```{r}
head(ASOIAF)
```

## Prepare and Validate Sex Codes

Many pedigree-based algorithms rely on biological sex for downstream calculationss and visualization. We use `checkSex()` to inspect the sex variable, repairing inconsistencies programmatically.


```{r}
df_got <- checkSex(ASOIAF,
  code_male = "M",
  code_female = "F",
  repair = TRUE)


# Find the IDs of Jon Snow and Daenerys Targaryen

jon_id <- df_got %>%
  filter(name == "Jon Snow") %>%
  pull(ID)

dany_id <- df_got %>%
  filter(name == "Daenerys Targaryen") %>%
  pull(ID)

```


```{r}
df_repaired <- checkParentIDs(df_got,
  addphantoms = TRUE,
  repair = TRUE,
  parentswithoutrow = FALSE,
  repairsex = FALSE
) %>% mutate(
  famID = 1,
  affected = case_when(
    ID %in% c(jon_id, dany_id, "365") ~ 1,
    TRUE ~ 0
  )
)
```


## Visualize the Pedigree with `plotPedigree()`


Here is the classic pedigree plot of the Targaryen family, with Jon Snow and Daenerys Targaryen highlighted in black The `plotPedigree()` function from {BGmisc} provides a quick way to visualize the pedigree structure.

```{r, message=FALSE, warning=FALSE}

plotPedigree(df_repaired, 
             affected = df_repaired$affected, 
             verbose = FALSE)

```


## Visualize the Pedigree with `ggPedigree()`

Here is the same pedigree, but using `ggPedigree()` from {ggpedigree}. This function provides a more flexible and customizable way to visualize pedigrees, allowing for easy integration with other `ggplot2` functions.

```{r, message=FALSE, warning=FALSE}


ggPedigree(df_repaired,  status_col = "affected",  
           personID_col = "ID",  
           config = list(unaffected =  0,
                         code_male = "M",
                         affected = 1,
                         affected_shape = 4,
                         point_size = 3,
                         spouse_segment_color = "pink",
    sibling_segment_color = "blue",
    parent_segment_color = "green",
    offspring_segment_color = "black",
                         
                         
                         ped_width=14))

```








```{r}
data("potter")
ggPedigree(potter,
  famID_col = "famID",
  personID_col = "personID",
  code_male = 1
)
```

`ggPedigree()` automatically:
1. reshapes the data by family (`ped2fam()`),

2. recodes sex (`recodeSex()`),

3. computes node layout (`calculateCoordinates()`), and

4. draws spouse, parent, sibling, and offspring links (`calculateConnections()`).

The result is a `ggplot2` object, so you can add any ggplot syntax to it. 


# Customizing the plot

All aesthetics reside in a config list; anything you omit falls back to the default.

```{r}
ggPedigree(
  potter,
  famID_col = "famID",
  personID_col = "personID",
  code_male = 1,
  config = list(
    spouse_segment_color = "pink",
    sibling_segment_color = "blue",
    parent_segment_color = "green",
    offspring_segment_color = "black"
  )
)
```

Because the result is just a ggplot object, regular layering applies:

```{r}
ggPedigree(potter,
  famID_col = "famID",
  personID_col = "personID",
  code_male = 1
) +
  theme_bw(base_size = 12)
```



# Alternative coding schemes

Pass the appropriate `code_male` when males are coded differently.


```{r}
data("hazard")

p <- ggPedigree(
  hazard %>% mutate(affected = as.factor(ifelse(affected == TRUE, "affected", "unaffected"))),
  famID_col = "famID",
  personID_col = "ID",
  code_male = 0,
  status_col = "affected"
)

p
```

# Multiple families in one graphic

Facet to separate distinct pedigrees:
```{r}
p +
  facet_wrap(~famID, scales = "free_x")
```

# Changing the layout

```{r}
p +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line        = element_line(colour = "black"),
    axis.text.x      = element_blank(),
    axis.text.y      = element_blank(),
    axis.ticks.x     = element_blank(),
    axis.ticks.y     = element_blank(),
    axis.title.x     = element_blank(),
    axis.title.y     = element_blank()
  ) + scale_color_viridis(
    discrete = TRUE,
    labels = c("Female", "Male", "Unknown")
  )
```
