---
title: "Plotting pedigrees with `ggPedigree()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting pedigrees with `ggPedigree()`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```



# Introduction

This vignette demonstrates pedigree visualisation with `ggPedigree()` from `ggpedigree`.
Internally, `ggPedigree()` leverages helpers from both {BGmisc} (ped2fam(), recodeSex()) and {ggpedigree} (`calculateCoordinates()`, `calculateConnections()`), and returns a standard `ggplot2` object, so any familiar ggplot syntax can be added afterward.

We illustrate usage with two bundled data sets:

- `potter` – extended wizarding family

- `hazard` – historical royalty


# Basic usage

```{r libraries, message=FALSE, warning=FALSE}
library(ggpedigree) # ggPedigree lives here
library(BGmisc) # helper utilities & example data
library(ggplot2) # ggplot2 for plotting
library(viridis) # viridis for color palettes
library(tidyverse) # for data wrangling
```

```{r}
data("potter")
ggPedigree(potter,
  famID = "famID",
  personID = "personID"
)
```

`ggPedigree()` automatically:

1. reshapes the data by family (`ped2fam()`),

2. recodes sex (`recodeSex()`),

3. computes node layout (`calculateCoordinates()`), and

4. draws spouse, parent, sibling, and offspring links (`calculateConnections()`).

The result is a `ggplot2` object, so you can add any ggplot syntax to it. 


# Customizing the plot

All aesthetics reside in a config list; anything you omit falls back to the default.

```{r}
ggPedigree(
  potter,
  famID = "famID",
  personID = "personID",
  config = list(
    code_male = 1,
    sex_color = FALSE,
    spouse_segment_color = "pink",
    sibling_segment_color = "blue",
    parent_segment_color = "green",
    offspring_segment_color = "black"
  )
)
```

Because the result is just a ggplot object, regular layering applies:

```{r}
ggPedigree(potter,
  famID = "famID",
  personID = "personID"
) +
  theme_bw(base_size = 12)
```



# Additional customization

You can also add affected/unaffected status to the plot. The `status_col` argument allows you to specify a column in your data frame that contains the affected/unaffected status of each individual. This column should be a factor or character vector, and `ggPedigree()` will automatically recode it to a factor if necessary.

When you add the `status_col` argument, `ggPedigree()` will automatically color the affected individuals in one color and the unaffected individuals in another. You can indicate what the values of the affected/unaffected status are in your data frame using the `affected` and `unaffected` arguments in the `config` list. By default, `ggPedigree()` will overlay the affected individuals with a different shape, but you can customize this as well.

```{r}
data("hazard")

p <- ggPedigree(
  hazard, # %>% mutate(affected = as.factor(ifelse(affected == TRUE, "affected", "uneffected"))),
  famID = "famID",
  personID = "ID",
  status_col = "affected",
  config = list(
    code_male = 0,
    sex_color = TRUE,
    affected = TRUE,
    unaffected = FALSE,
    affected_shape = 4
  )
)

p
```

If you set sex_color to FALSE, the affected values will be filled with the default color palette.

```{r}
ggPedigree(
  hazard,
  famID = "famID",
  personID = "ID",
  status_col = "affected",
  config = list(
    code_male = 0,
    sex_color = FALSE,
    affected = TRUE,
    unaffected = FALSE
  )
)
```

# Multiple families in one graphic

If you have multiple families in your data frame, you can use the `facet_wrap()` function from `ggplot2` to create separate plots for each family. This is useful if you want to compare the pedigrees of different families side by side.

Facet to separate distinct pedigrees, note that you can use `scales = "free_x"` to allow each family to have its own x-axis scale.

```{r}
p +
  facet_wrap(~famID, scales = "free_x")
```

# Changing the layout

```{r}
p +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line        = element_line(colour = "black"),
    axis.text.x      = element_blank(),
    axis.text.y      = element_blank(),
    axis.ticks.x     = element_blank(),
    axis.ticks.y     = element_blank(),
    axis.title.x     = element_blank(),
    axis.title.y     = element_blank()
  ) + scale_color_viridis(
    discrete = TRUE,
    labels = c("Female", "Male", "Unknown")
  )
```

# More complex examples

```{r message=FALSE, warning=FALSE}
library(BGmisc) # helper utilities & example data

data("inbreeding")

df <- inbreeding 

#df  <- dplyr::filter(df, famID %in% c(5, 7))


p <- ggPedigree(
  df,
  famID = "famID",
  personID = "ID",
  status_col = "proband",
#  debug = TRUE,
  config = list(
    code_male = 0,
    sex_color = F,
  #  label_method = "geom_text",
    affected = TRUE,
    unaffected = FALSE,
    generation_height = 2,
    generation_width = 1,
    affected_shape = 4,
    spouse_segment_color = "pink",
    sibling_segment_color = "blue",
    parent_segment_color = "green",
    offspring_segment_color = "black"
  )
) 

# p$connections%>%filter(personID     ==60) %>% nrow()
# p$connections%>%filter(personID     ==66) %>% unique()
# p$connections%>%filter(personID     ==65) %>% unique()

# p$connections%>%filter(personID  >=61 &
#       personID  <62 ) %>% unique()

p  + facet_wrap(~famID, scales= "free") #+ scale_color_viridis(
 #   discrete = TRUE,
 #   labels = c("TRUE", "FALSE")
#  )  + theme_bw(base_size = 14)  +  guides(colour="none", shape="none")
```

# Power paper
```{r}
library(tibble)

pedigree_df <- tribble(
  ~personID, ~momID, ~dadID, ~sex, ~famID,
  10011,     NA,     NA,     0,    1,
  10012,     NA,     NA,     1,    1,
  10021,     NA,     NA,     1,    1,
  10022,  10011,  10012,     1,    1,
  10023,  10011,  10012,     0,    1,
  10024,     NA,     NA,     0,    1,
  10025,     NA,     NA,     0,    1,
  10026,  10011,  10012,     0,    1,
  10027,  10011,  10012,     1,    1,
  10031,  10023,  10021,     0,    1,
  10032,  10023,  10021,     1,    1,
  10033,  10023,  10021,     1,    1,
  10034,  10023,  10021,     1,    1,
  10035,  10023,  10021,     0,    1,
  10036,  10024,  10022,     1,    1,
  10037,  10024,  10022,     0,    1,
  10038,  10025,  10027,     1,    1,
  10039,  10025,  10027,     0,    1,
  10310, 10025,  10027,     1,    1,
  10311, 10025,  10027,     1,    1,
  10312, 10025,  10027,     0,    1,
  10011,     NA,     NA,     0,    2,
  10012,     NA,     NA,     1,    2,
  10021,     NA,     NA,     0,    2,
  10022,  10011,  10012,     0,    2,
  10023,  10011,  10012,     1,    2,
  10024,  10011,  10012,     1,    2,
  10025,     NA,     NA,     1,    2,
  10026,  10011,  10012,     0,    2,
  10027,     NA,     NA,     1,    2,
  10031,  10021,  10023,     1,    2,
  10032,  10021,  10023,     0,    2,
  10033,  10021,  10023,     1,    2,
  10034,  10022,  10025,     0,    2,
  10035,  10022,  10025,     0,    2,
  10036,  10022,  10025,     1,    2,
  100310, 10022,  10025,     1,    2,
  10037,  10026,  10027,     0,    2,
  10038,  10026,  10027,     0,    2,
  10039,  10026,  10027,     0,    2,
  100311, 10026,  10027,     1,    2,
  100312, 10026,  10027,     1,    2
) %>% mutate (proband = TRUE)

#pedigree_df <- recodeSex(pedigree_df,code_male = 1, recode_male = "M")
pedigree_df$personID[pedigree_df$famID == 1] <- pedigree_df$personID[pedigree_df$famID == 1]-10000
pedigree_df$momID[pedigree_df$famID == 1] <- pedigree_df$momID[pedigree_df$famID == 1]-10000
pedigree_df$dadID[pedigree_df$famID == 1] <- pedigree_df$dadID[pedigree_df$famID == 1]-10000



p <- ggPedigree(
  pedigree_df,
  famID = "famID",
  personID = "personID",
  status_col = "proband",
#  debug = TRUE,
  config = list(
    code_male = 1,
    sex_color = F,
    apply_default_scales = FALSE,
   label_method = "geom_text",
 #   affected = TRUE,
 #   unaffected = FALSE,
    generation_height = 1,
    generation_width = 1,
    affected_shape = 4,
    spouse_segment_color = "black",
    sibling_segment_color = "black",
    parent_segment_color = "black",
    offspring_segment_color = "black"
  )
)

p + scale_shape_manual(
      values = c(16, 15, 15),
      labels =  c("Female", "Male", "Unknown")
    ) +
  guides(colour="none", shape="none")
 # guides(colour="none", shape="none") + 
#facet_wrap(~famID, scales= "free") 


```
