---
title: "Visualizing Relatedness Matrices with `ggRelatednessMatrix()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing Relatedness Matrices with `ggRelatednessMatrix()`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages({
  library(BGmisc)
  library(ggplot2)
  library(readxl)
  library(dplyr)
  library(reshape2)
})

```

# Introduction

This vignette demonstrates how to calculate and visualize relatedness matrices for animal pedigrees using functions from the BGmisc and ggpedigree packages.
We focus on Kluane Red Squirrel Project data, but the workflow generalizes to any pedigree dataset.

We will use the new ggRelatednessMatrix function to create ggplot2-based heatmaps of relatedness matrices, inspired by the approaches in the original squirrel population analyses.

## Data Preparation

First, load pedigree data (here, as in the original vignette, from Excel):

```{r}

# Load the example data
data("redsquirrels")


```


# Calculating Relatedness Matrices

Here we show how to compute additive genetic and mitochondrial relatedness matrices for a target family, as in the red squirrel analysis.

```{r}

sumped <- summarizePedigrees(ds, famID = "famID", personID = "personID",  nbiggest = 5)


# Set target family for visualization
fam_filter <- sumped$biggest_families$famID[3]

# Filter for the largest family, recode sex if needed
ped_filtered <- ds %>%
  recodeSex(code_female = "F") %>%
  filter(famID == fam_filter)

# Calculate relatedness matrices
add_mat <- ped2add(ped_filtered, isChild_method = "partialparent", sparse = FALSE)
mit_mat <- ped2mit(ped_filtered, isChild_method = "partialparent", sparse = FALSE)
```


Visualizing Relatedness Matrices
The ggRelatednessMatrix function plots a relatedness matrix as a heatmap, with options for clustering and color customization. This approach mirrors the style of plots produced in the original red squirrel data analysis.


# Plot additive genetic relatedness

```{r}
p_add <- ggRelatednessMatrix(
  add_mat,
  config = list(
    color_palette = c("white", "orange", "red"),
    scale_midpoint = 0.55,
    cluster = TRUE,
    title = "Additive Genetic Relatedness",
    text_size = 15
  )
)
p_add
```


# Plot mitochondrial relatedness

```{r}
p_mit <- ggRelatednessMatrix(
  mit_mat,
  config = list(
    color_palette = c("white", "skyblue", "darkblue"),
    scale_midpoint = 0.55,
    cluster = TRUE,
    title = "Mitochondrial Relatedness",
    text_size = 6
  )
)
p_mit
```

Customization Options
You can adjust clustering, color scales, labels, and more using the config list. For example, to turn off clustering:

```{r}
p_add_noclust <- ggRelatednessMatrix(
  add_mat,
  config = list(cluster = FALSE, title = "Additive Relatedness (No Clustering)")
)
p_add_noclust
```

Comparison to Base R and Other Approaches
For comparison, here is how the same matrix would be plotted in base R, as in earlier squirrel analyses:

```{r}
if (requireNamespace("corrplot", quietly = TRUE)) {
  corrplot::corrplot(
    as.matrix(add_mat),
    method = "color",
    type = "lower",
    col.lim = c(0, 1.25),
    is.corr = FALSE,
    title = "Additive Relatedness (Base R)",
    order = "hclust",
    col = corrplot::COL1('Reds', 100)
  )
}
```

# Conclusions

The ggRelatednessMatrix function provides a flexible and visually appealing way to explore patterns of genetic and mitochondrial relatedness within families. This workflow integrates seamlessly with tools from BGmisc and ggpedigree, supporting reproducible, publication-quality visualization for quantitative genetic analysis.

