---
title: "Extended Plotting pedigrees with ggPedigree()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extended Plotting pedigrees with ggPedigree()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```
```{r libraries, message=FALSE, warning=FALSE}
library(ggpedigree) # ggPedigree lives here
library(BGmisc) # helper utilities & example data
library(ggplot2) # ggplot2 for plotting
library(viridis) # viridis for color palettes
library(tidyverse) # for data wrangling
```


# Constructing Custom Pedigrees for Publication

Here we demonstrate how to create a custom pedigree using the `ggpedigree package`. The data shown here were generated using the `simulatePedigree()` function from the {BGmisc} package, which is the parent package to {ggpedigree}. These simulated pedigrees were used in a study evaluating statistical power and estimation bias for a variance decomposition model that includes mitochondrial DNA (mtDNA) effects.

The simulation generated thousands of extended pedigree structures varying in depth, sibship size, mating structure, and maternal lineage overlap. The example below shows one of the simulated pedigrees and is the version included in the final manuscript:

> Detecting mtDNA effects with an Extended Pedigree Model: An Analysis of Statistical Power and Estimation Bias
> Xuanyu Lyu, S. Alexandra Burt, Michael D. Hunter, Rachel Good, Sarah L. Carroll, S. Mason Garrison
> Preprint available at: https://doi.org/10.1101/2024.12.19.629449

The structure includes multiple generations, sibling sets, and overlapping parental lineages, and was chosen to illustrate the complexity of the simulated pedigrees used in the power study.

## Preparing the data

Each row represents one individual. Variables include `personID`, `momID`, `dadID`, `sex`, and `famID`. The proband variable is included to demonstrate status overlays. For plotting, we normalize identifiers in family 1 to avoid ID collisions across families. 

<details>
<summary>Click to expand pedigree setup</summary>

```{r, message=FALSE, warning=FALSE}
library(tibble)
library(dplyr)
pedigree_df <- tribble(
  ~personID, ~momID, ~dadID, ~sex, ~famID,
  10011, NA, NA, 0, 1,
  10012, NA, NA, 1, 1,
  10021, NA, NA, 1, 1,
  10022, 10011, 10012, 1, 1,
  10023, 10011, 10012, 0, 1,
  10024, NA, NA, 0, 1,
  10025, NA, NA, 0, 1,
  10026, 10011, 10012, 0, 1,
  10027, 10011, 10012, 1, 1,
  10031, 10023, 10021, 0, 1,
  10032, 10023, 10021, 1, 1,
  10033, 10023, 10021, 1, 1,
  10034, 10023, 10021, 1, 1,
  10035, 10023, 10021, 0, 1,
  10036, 10024, 10022, 1, 1,
  10037, 10024, 10022, 0, 1,
  10038, 10025, 10027, 1, 1,
  10039, 10025, 10027, 0, 1,
  10310, 10025, 10027, 1, 1,
  10311, 10025, 10027, 1, 1,
  10312, 10025, 10027, 0, 1,
  10011, NA, NA, 0, 2,
  10012, NA, NA, 1, 2,
  10021, NA, NA, 0, 2,
  10022, 10011, 10012, 0, 2,
  10023, 10011, 10012, 1, 2,
  10024, 10011, 10012, 1, 2,
  10025, NA, NA, 1, 2,
  10026, 10011, 10012, 0, 2,
  10027, NA, NA, 1, 2,
  10031, 10021, 10023, 1, 2,
  10032, 10021, 10023, 0, 2,
  10033, 10021, 10023, 1, 2,
  10034, 10022, 10025, 0, 2,
  10035, 10022, 10025, 0, 2,
  10036, 10022, 10025, 1, 2,
  10310, 10022, 10025, 1, 2,
  10037, 10026, 10027, 0, 2,
  10038, 10026, 10027, 0, 2,
  10039, 10026, 10027, 0, 2,
  10311, 10026, 10027, 1, 2,
  10312, 10026, 10027, 1, 2
) %>%
  mutate(
    cleanpersonID = personID - 10000,
    personID = ifelse(famID == 1, personID - 10000, personID),
    momID = ifelse(famID == 1 & !is.na(momID), momID - 10000, momID),
    dadID = ifelse(famID == 1 & !is.na(dadID), dadID - 10000, dadID),
    proband = case_when(
      personID %in% c(11, 22, 23, 26, 27, 31, 32, 33, 34, 35) ~ TRUE,
      personID %in% c(
        10011, 10022, 10022, 10023, 10024, 10026,
        10034, 10035, 10036, 10310,
        10037, 10038, 10039, 10311,
        10312
      ) ~ TRUE,
      TRUE ~ FALSE
    )
  )


df_fig1 <- tribble(
  ~personID, ~momID, ~dadID, ~sex, ~famID,
  10011, NA, NA, 0, 1,
  10012, NA, NA, 1, 1,
  10021, NA, NA, 1, 1,
  10022, 10011, 10012, 1, 1,
  10023, 10011, 10012, 0, 1,
  10024, NA, NA, 0, 1,
  10025, 10011, 10012, 0, 1,
  10027, NA, NA, 1, 1,
  10031, 10023, 10021, 0, 1,
  10032, 10023, 10021, 1, 1,
  10035, 10023, 10021, 0, 1,
  10036, 10024, 10022, 1, 1,
  10037, 10024, 10022, 0, 1,
  10038, 10025, 10027, 1, 1
) %>%
  mutate(
    proband = case_when(
      personID %in% c(10011, 10022, 10023, 10025, 10031, 10032, 10035, 10038) ~ TRUE,
      TRUE ~ FALSE
    ),
    mtdnaline2 = case_when(
      personID %in% c(10024, 10036, 10037) ~ TRUE,
      TRUE ~ FALSE
    ),
  )
```
</details>


## Plotting the pedigree
```{r, dimensions = c(6, 4)}
fig1 <- ggPedigree(
  df_fig1,
  famID = "famID",
  personID = "personID",
  status_column = "proband",
  debug = TRUE,
  config = list(
    code_male = 1,
    sex_color_include = FALSE,
    apply_default_scales = FALSE,
    label_method = "geom_text",
    label_col = "personID",
    point_size = 5,
    outline_include = TRUE,
    status_code_affected = TRUE,
    status_code_unaffected = FALSE,
    generation_height = 1,
    generation_width = 1,
    status_affected_shape = 4,
    segment_spouse_color = "black",
    segment_sibling_color = "black",
    segment_parent_color = "black",
    segment_offspring_color = "black",
    outline_multiplier = 1.25,
    segment_linewidth = .5
  )
)
# fig1

fig1$plot + geom_point(aes(x = x_pos, y = y_pos),
  color = "cornflowerblue", size = 2,
  data = fig1$data %>% dplyr::filter(mtdnaline2 == TRUE)
) +
  scale_shape_manual(
    values = c(16, 15, 14),
    labels = c("Female", "Male", "Unknown")
  ) +
  guides(shape = "none") + scale_color_manual(
    values = c("pink", "white")
  ) +

  # discrete = TRUE,
  # labels = c("TRUE", "FALSE"),
  #  name = ""
  # ) +
  theme(
    strip.text = element_blank(),
    legend.position = "none"
  )
```



```{r}
p2 <- ggPedigree(
  pedigree_df,
  famID = "famID",
  personID = "personID",
  status_column = "proband",
  #  debug = TRUE,
  config = list(
    code_male = 1,
    sex_color_include = FALSE,
    apply_default_scales = FALSE,
    label_method = "geom_text",
    label_include = TRUE,
    label_column = "cleanpersonID",
    status_code_affected = TRUE,
    status_code_unaffected = FALSE,
    generation_height = 1,
    generation_width = 1,
    status_affected_shape = 4,
    segment_spouse_color = "black",
    segment_sibling_color = "black",
    segment_parent_color = "black",
    segment_offspring_color = "black"
  )
)
```


We finish by adjusting the legend and shape scale for visual clarity:

```{r, message=FALSE, warning=FALSE,dimensions = c(6, 4)}
p2 + scale_shape_manual(
  values = c(16, 15, 14),
  labels = c("Female", "Male", "Unknown")
) +
  guides(shape = "none") + scale_color_viridis(
    discrete = TRUE,
    labels = c("TRUE", "FALSE"),
    name = "Founding MtDNA Line"
  ) +
  facet_wrap(~famID, scales = "free", shrink = TRUE) +
  theme(
    strip.text = element_blank(),
    legend.position = "bottom"
  )
```

# More Complex Pedigree Plots with ggPedigree

In this section, we demonstrate how to create a more complex pedigree plot with multiple families. We use the `inbreeding` dataset from the `BGmisc` package, which contains several multigenerational pedigrees with consanguinity. Note that in these plots that some individuals may appear in multiple places within the pedigree. This is common in large pedigrees, especially when there are overlapping generations or multiple marriages. Here the colors are set to be the same for all segments, except for self-loops, which are colored purple.

```{r self-loops, message=FALSE, warning=FALSE}
library(BGmisc) # helper utilities & example data

data("inbreeding")


df <- inbreeding # multigenerational pedigree with consanguinity


# df  <- dplyr::filter(df, famID %in% c(5, 7))


p <- ggPedigree(
  df,
  famID = "famID",
  personID = "ID",
  status_column = "proband",
  #  debug = TRUE,
  config = list(
    code_male = 0,
    sex_color_include = FALSE,
    status_code_affected = TRUE,
    status_code_unaffected = FALSE,
    generation_height = 1,
    point_size = 2,
    generation_width = 1,
    status_affected_shape = 4,
    segment_self_color = "purple"
  )
) 

p + facet_wrap(~famID, scales = "free") #+ scale_color_viridis(
#   discrete = TRUE,
#   labels = c("TRUE", "FALSE")
#  )  + theme_bw(base_size = 14)  +  guides(colour="none", shape="none")
```


